### 2013-03-17 doc/sosp13/memory.tex
lots of edits to memory section.

Applications written for Grappa utilize two forms of memory: local and global.
Local memory is local to a single core \comment{core or processor/node?} in
the system. Accesses occur through conventional pointers. The compiler emits
an access and the memory is manipulated directly. Applications use local
accesses for a number of things in Grappa: the stack associated with a task,
accesses to localized global memory in caches (see below), and accesses to
debugging infrastructure that is local to each system node. Local pointers
cannot access memory on other cores, and are valid only on their home core.

@@

\begin{figure}[htbp]
  \begin{center}
    \begin{minipage}{\columnwidth}
	\small
	\begin{tabular}{l}
      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
      	\texttt{\scriptsize global\_free( global\_address )} \\ 
      	Allocation in the global heap \\ \\
      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ 
      	Memory operations at the home core of a global address \\ \\
      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ 
		Perform cache operations to acquire/release global data.  \\
		Acquire copies all data to local node and returns a pointer. \\ 	
		Release optionally copies data back to global memory. \\
	\end{tabular}
      \caption{\label{fig:accessing-memory} Grappa API for memory accesses. \TODO{Update table with new calls. And async calls for delegates.}}     \end{minipage}
  \end{center}
\end{figure}

@@

\paragraph{Global memory access} There are two general approaches
Grappa applications use to {\emph access} global memory. When the
programmer expects a computation on shared data to have spatial locality
to exploit, {\em cache} operations may be used. When there is no
locality to exploit, {\em delegate} operations are used.

\textit{Explicit caching.} Grappa provides an API to fetch a global pointer of
any length and return a local pointer to a cached copy of the global memory.
Grappa cache operations have the usual read-only and read-write variants,
along with a write-only variant used to initialize data structures. Languages
for distributed shared memory systems have done optimizations to achieve a
similar goal. For example, the UPC compiler coalesces struct and array
accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's
message vectorization hoists small messages out of a loop
\cite{FortranD:1992}. Caching in Grappa additionally provides a mechanism for
exploiting temporal locality when available by operating on the data locally.

Under the hood, Grappa performs the mechanics of gathering chunks of data from
multiple system nodes and presenting a conventional appearing linear block of
memory as a local pointer into a cache. The strategy employed is to issue all
the constituent requests of the desired block (as Active Messages) and then
yield until all responses have occurred. Currently, Grappa caches are
\emph{not} kept coherent automatically. If that is desired, the programmer to
is responsible for pushing the updated block back to its home node. Given that
synchronization is explicit in Grappa, we don't expect this to be a
significant burden (programmers can push updates before a relevant
synchronization operation).

% Future work will develop a software directory based coherence scheme to
% simplify consistent access to global data.

\textit{Delegate operations.} When the access pattern has low-locality,
it is more efficient to modify the data on its home core rather than
bringing a copy to the requesting core and returning it after
modification. Delegate operations provide this capability. Applications
can dispatch computation to be performed on individual machine-word
sized chunks of global memory to the memory system itself (e.g.,
\emph{fetch-and-add}).  Delegate operations, proposed in
\cite{Nelson:hotpar11} and \cite{delegated:oopsla11}, are used to implement synchronization in Grappa.

Delegate operations are always executed at the home core of their
address, and while arbitrary memory operations can be delegated, we
restrict the use of delegate operations in three ways to make them more
useful for synchronization. First, we limit each task to one outstanding
delegate operation to avoid the possibility of reordering in the
network. Second, we limit delegate operations to operate on objects in
the 2D address space or objects that fit in a single block of the linear
address space so they can be satisfied with a single network request.
Finally, no context switches are allowed while the data is being
modified. Given these restrictions, we can ensure that delegate
operations for the same address from multiple requesters are always
serialized through a single core in the system, providing atomic
semantics without using actual atomic operations (and thus avoiding their typical high cost). 
Figure~\ref{fig:delegate-cache} depicts an example of how delegate and
cache operations interact.

\TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}

@@

\paragraph{Memory consistency model discussion} As mentioned earlier, all
synchronization operations are done via delegate operations. Since they all
execute in their home node, they are guaranteed to be serialized , with their
updates visible to all cores across the system in the same order. Also, tasks
are restricted to a single outstanding delegate operation at a time, and
therefore are not subject to potential request reordering. Consequently, all
synchronization operations execute in program order and are made visible in
the same order to all cores in the system. These properties are sufficient to
guarantee a memory model that offers sequential consistency for data-race-free
programs~\cite{AdveHill1990} (all accesses to shared data are separated by
synchronization). This is the memory model that underpins C/C++~\cite{N2480,N2800}.

Note, however, that if the application code uses explicit caching on shared
data, all updates must be sent back to the home node before the
synchronization operation that protects the data is performed.

>>>>

Applications written for Grappa utilize two forms of memory: local and global.
Local memory is local to a single core within a node in the system. Accesses
occur through conventional pointers. The compiler emits an access and the
memory is manipulated directly. Applications use local accesses for a number
of things in Grappa: the stack associated with a task, accesses to localized
global memory in caches (see below), and accesses to debugging infrastructure
that is local to each system node. Local pointers cannot access memory on
other cores, and are valid only on their home core.

@@

\begin{figure}[htbp]
  \begin{center}
    \begin{minipage}{\columnwidth}
	\small
	\begin{tabular}{l}
      	\texttt{\scriptsize global\_address global\_malloc( size )} \\
      	\texttt{\scriptsize global\_free( global\_address )} \\ 
      	Allocation in the global heap \\ \\
      	\texttt{\scriptsize delegate\_read( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_read\_async( global\_address, local\_var )}  \\
      	\texttt{\scriptsize delegate\_write( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_write\_async( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_cas( global\_address, local\_var )} \\
      	\texttt{\scriptsize delegate\_fetch\_inc( global\_address, local\_var )} \\ 
      	Memory operations at the home core of a global address \\ \\
      	\texttt{\scriptsize cache\_acquire( global\_address, local\_buf, \{RO,RW,WO\})} \\
      	\texttt{\scriptsize cache\_release( global\_address, local\_buf )} \\ 
		Perform cache operations to acquire/release global data.  \\
		Acquire copies all data to local node and returns a pointer. \\ 	
		Release optionally copies data back to global memory. \\
	\end{tabular}
      \caption{\label{fig:accessing-memory} Grappa API for memory accesses. \TODO{Update table with new calls. And async calls for delegates.}}     \end{minipage}
  \end{center}
\end{figure}

@@

\paragraph{Global memory access} There are two general methods Grappa
applications use to {\em access} global memory. One method is to use {\em
delegate} operations, which are short memory accesses performed at the memory
location's home node. This method is used when there is no locality in the
access. The other method is to do {\em explicit caching} of memory regions,
used when the programmer expects that the associated computation has data
locality. Below we explain each in detail.

\vspace{1ex} \textit{Delegate operations.} When the data access pattern has
low-locality, it is more efficient to modify the data on its home core rather
than bringing a copy to the requesting core and returning it after
modification. Delegate operations~\cite{Nelson:hotpar11, delegated:oopsla11}
provide this capability. Applications can dispatch computation to be performed
on individual machine-word sized chunks of global memory to the memory system
itself. Delegates are general, so we use them to perform simple
\emph{read}/\emph{write} operations to global memory, as well as more complex \emph{read-modify-write} operations (e.g.,\emph{fetch-and-add}).

Delegate operations are \emph{always} executed at the home core of their
address. The remote operation may not perform any operations that could cause
a context switch; this ensures any modifications are atomic. We limit delegate
operations to operate on objects in the 2D address space or objects that fit
in a single block of the linear address space so they can be satisfied with a
single network request. Given these restrictions, we can ensure that delegate
operations for the same address from multiple requesters are always serialized
through a single core in the system, providing atomic semantics without using
actual atomic operations (and thus avoiding their typical high cost).

% We restrict the use of delegate operations in order to make them also
% useful for synchronization operations.

Delegate operations can be either {\em synchronous} or {\em asynchronous}.
With synchronous operations, the task issuing the delegate call blocks until
the delegate operation completes, which is necessary, for example, to ensure
that synchronization has finished before continuing. Synchronization in Grappa
is implemented using synchronous delegate operations. Often remote data
accesses do not need to be performed synchronously. In order to avoid
unnecessary waiting, we support asynchronous delegate operations. For reads,
we support a ``futures''-like mechanism. Tasks may issue delegate reads in
parallel and block on the ``promises'' returned by asynchronous delegate
invocations. Delegate write operations may also be performed asynchronously,
and since they do not return data, they need a mechanism to detect completion.
Grappa provides a \texttt{GlobalCompletionEvent} object, which keeps track of
outstanding asynchronous operations of the local core. The code and can probe
the state of this object to determine when all outstanding requests are
complete. This is necessary to, say, make sure all outstanding operations are
complete before executing a synchronization operation.

% \TODO{Talk about synchronous and asynchronous delegate operations. Talk about global completion object. Should we have a whole section on synchronization?}
% \TODO{If we do want a section on GlobalCompletionEvent:
%   \begin{itemize}
%     \item many tasks and operations in the applications we've implemented do not need immediate consistency
%     \item guaranteeing ordering requires blocking a task. this is unavoidable if the task relies on the returned value, or if one task is dependent on others
%     \item however, much of the applications we've implemented (and other big-data things) are data-parallel, so tasks within a data-parallel loop can run in any order
%     \item in addition, often these data-parallel tasks will update values that aren't used until the next data-parallel loop
%     \item this allows us to free up workers/stacks (running out of stacks causes  latency-sensitivity because we can't continue until tasks complete)
%     \item must guarantee all operations that were supposed to happen in the loop complete before continuing, including tasks and delegate operations
%     \item \emph{motivating example:} write to array in one loop, use updated values in next loop (could also mention the ``PushBuffer'' example where we buffer queue appends because the order in the queue doesn't matter)
%     \item GlobalCompletionEvent is complicated by task-stealing and the fact that delegate operations complete on a different core than they started, forcing us to use a ``cancelable barrier''-like approach (possible citation: Ravichandran, et al. ``Work stealing for multi-core HPC Clusters'')
%     \item interface to GlobalCompletionEvent:
%       \begin{itemize}
%         \item \texttt{gce.enroll(n)}: enroll n tasks/operations to be done, when goes from 0 to 1, leave cancelable barrier
%         \item \texttt{gce.complete(n)}: complete tasks, when it goes to zero, enter global cancelable barrier
%         \item \texttt{gce.wait()}: block until all cores have no work
%       \end{itemize}
%     \item to prevent early wakeups, stolen tasks must send a completion message back to the core they were started on
%   \end{itemize}
% }

\vspace{1ex}
\textit{Explicit caching.} Grappa provides an API to fetch a global pointer of
any length and return a local pointer to a cached copy of the global memory.
Grappa cache operations have the usual read-only and read-write variants,
along with a write-only variant used to initialize data structures. Languages
for distributed shared memory systems have done optimizations to achieve a
similar goal. For example, the UPC compiler coalesces struct and array
accesses into remote get/put \cite{Chen:2005}, and Fortran D compiler's
message vectorization hoists small messages out of a loop
\cite{FortranD:1992}. Caching in Grappa additionally provides a mechanism for
exploiting temporal locality when available by operating on the data locally.

Under the hood, Grappa performs the mechanics of gathering chunks of data from
multiple system nodes and presenting a conventional appearing linear block of
memory as a local pointer into a cache. The strategy employed is to issue all
the constituent requests of the desired block (as Active Messages) and then
yield until all responses have occurred. Currently, Grappa caches are
\emph{not} kept coherent automatically. If that is desired, the programmer is
responsible for ensuring sufficient synchronization (via delegate operations)
occurs before acquiring or releasing cache objects.

Figure~\ref{fig:delegate-cache} depicts an example how explicit caching and
delegate operations work and how they interact.

@@

\paragraph{Memory consistency model discussion} As mentioned earlier, all
synchronization operations are done via delegate operations. Since they all
execute in their home node, they are guaranteed to be serialized, with their
updates visible to all cores across the system in the same order. Also, tasks
are restricted to a single outstanding synchronous delegate operation at a
time, which is how synchronization is done. Therefore, synchronization operations are not subject to potential reordering. 
% \TODO{I think we can support multiple delegates in parallel from a task as
% long as we block on them before counting on them being complete. see my
% description above about the \emph{future/promise} delegates we have now. Not
% sure if there's a strong example of when this is useful for synchronization
% (though it's definitely useful for reads/writes), acquiring multiple locks
% doesn't work because we have to acquire locks in order to prevent deadlock.}
Consequently, all synchronization operations execute in program order and are
made visible in the same order to all cores in the system. These properties
are sufficient to guarantee a memory model that offers sequential consistency
for data-race-free programs~\cite{AdveHill1990} (all accesses to shared data
are separated by synchronization). This is the memory model that underpins
C/C++~\cite{N2480,N2800}.

Note, however, that if the application code uses explicit caching or
asynchronous delegates to access shared data, all updates must be sent back to
the home node before the synchronization operation that protects the data is
performed. This is done using release operations on cached regions and using the \texttt{GlobalCompletionEvent} object to determine that asynchronous delegates have been fully executed.

% \TODO{Not sure this warrants its own paragraph, seems like the explicit caches are just the same as reading and then writing using delegates.}
